\textbf{Seasonal Dynamics of Epiphytic Microbial Communities on Marine Macrophyte Surfaces}

### 2020-03-17 submission/manuscript.tex
Analysis rerun.

Sequencing of the 16S rRNA V4 region yielded a total of 1.8 million
sequences after quality curation and exclusion of eukaryotic,
chloroplast, mitochondrial and no relative sequences
(\autoref{supp-nseq_notus}). A total of 35 samples originating from
epiphytic archaeal and bacterial communities associated with surfaces of
the seagrass \emph{C. nodosa} and macroalga \emph{C. cylindracea} were
analyzed. In addition, 18 samples (one of the samples was sequenced two
times) originating from picoplankton archaeal and bacterial communities
in the ambient seawater were also processed for comparison. The number
of reads per sample ranged between 8,410 and 77,465 sequences
(\autoref{supp-nseq_notus}). Even when the highest sequencing effort was
applied the rarefaction curves did not level off that is a common
observation in high-throughput 16S rRNA amplicon sequencing approaches
(\autoref{supp-rarefaction}). Following quality curation and exclusion
of sequences mentioned before reads were clustered into 28,729 different
OTUs at a similarity level of 97 \si{\percent}. Reads numbers were
normalized to the minimum number of sequences, 8,410
(\autoref{supp-nseq_notus}), through rarefaction resulting in 17,019
different OTUs that ranged from 355 to 2,018 OTUs per sample
(\autoref{supp-calculators}). To determine seasonal changes of richness
and diversity the Observed Number of OTUs, Chao1, ACE, Exponential
Shannon (Jost, 2006) and Inverse Simpson were calculated after
normalization through rarefaction. Generally, richness estimators and
diversity indices showed similar trends. On average, higher values were
found for \emph{C. cylindracea} (invaded {[}Number of OTUs, 1,690.5 ±
152.3 OTUs{]} and noninvaded {[}Number of OTUs, 1,733.8 ± 163.5
OTUs{]}), middle vales for \emph{C. nodosa} (Number of OTUs, 1,066.9 ±
224.5 OTUs) and lower values for picoplankton communities in the ambient
seawater (Number of OTUs, 526 ± 145.6 OTUs)
(\autoref{supp-calculators}). Seasonal changes did not show such large
dissimilarities. \emph{C. nodosa} communities showed a slow increase
towards the end of the study, while \emph{C. cylindracea} (invaded and
noninvaded) communities were characterized by slightly larger values in
Spring and Summer in comparison to Autumn and Winter
(\autoref{supp-calculators}).

To determine the proportion of shared archaeal and bacterial OTUs and
communities sampled in different environments the Jaccard's Similarity
Coefficient on presence-absence data and Bray-Curtis Similarity
Coefficient were, respectively, calculated. Coefficients were determined
after normalization through rarefaction and binning of samples from a
particular environment. The highest proportion of shared OTUs and
community was found between invaded and noninvaded \emph{C. cylindracea}
(Jaccard, 0.35; Bray-Curtis, 0.77), while lower shared values were
calculated between seawater and epiphytic communities
(\autoref{matrix}). Shared proportion between \emph{C. nodosa} and
\emph{C. cylindracea} were approximately in the middle between these two
extremes. To assess seasonal changes in the proportion of shared OTUs
and communities the Jaccard's and Bray-Curtis Similarity Coefficients
were calculated between consecutive sampling points (\autoref{shared}).
Both coefficients showed similar trends. Temporal proportional changes
were more pronounced for seawater in comparison to \emph{C. nodosa} and
especially \emph{C. cylindracea} associated communities
(\autoref{shared}). In addition, only 0.4 -- 1.0 \si{\percent} of OTUs
from each surface associated community were found at every time point.
These OTUs made also a high proportion of total sequences (39.9 -- 52.8
\si{\percent}). To further disentangle the environmental and seasonal
community dissimilarity a Principal Coordinates Analysis (PCoA) based on
Bray-Curtis distances and OTU abundances was applied. It showed a clear
separation between planktonic and surface associated communities
(\autoref{pcoa}). In addition, a separation of epiphytic bacterial and
archaeal communities based on host species was determined. This
separation was further supported by ANOSIM (R = 0.96, \emph{p}
\textless{} 0.001). Seasonal changes of \emph{C. nodosa} associated
communities indicated a separation between Spring, Summer and
Autumn/Winter samples (ANOSIM, R = 0.54, \emph{p} \textless{} 0.01),
while communities from the surfaces of \emph{C. cylindracea} indicated a
non so strongly supported, as in previous cases, separation between
Summer and Autumn/Winter/Spring samples (ANOSIM, R = 0.31, \emph{p}
\textless{} 0.05) (\autoref{pcoa}).

@@

\emph{Cyanobacteria} related sequences were comprising, on average, 5.5
± 4.4 \si{\percent} of total sequences (\autoref{cyano}). Higher
proportions were found for \emph{C. nodosa} (16.4 ± 5.3 \si{\percent})
and \emph{C. cylindracea} (invaded {[} (7.7 ± 3.9 \si{\percent}){]} and
noninvaded {[} (7.8 ± 2.4 \si{\percent}){]}) associated communities in
autumn and for seawater communities in winter (8.8 ± 7.4 \si{\percent}).
Large taxonomic differences between surface associated and seawater
cyanobacterial communities were observed. Seawater communities were
mainly comprised of \emph{Cyanobium} and \emph{Synechococcus}, while
surface associated communities were consisted of \emph{Pleurocapsa} and
sequences without known relatives within \emph{Cyanobacteriia}
(\autoref{cyano}). In addition, seasonal changes in surface associated
communities were observed with \emph{Pleurocapsa} and no relative
\emph{Cyanobacteriia} comprising larger proportions in autumn and winter
and \emph{Acrophormium}, \emph{Phormidesmis} and no relative
\emph{Nodosilineaceae} in spring and summer (\autoref{cyano}).

@@

Sequences related to \emph{Gammaproteobacteria} were comprising, on
average, 18.7 ± 3.9 \si{\percent} of all sequences (\autoref{gamma}).
Similarly to previous taxa, large taxonomic differences between seawater
and surface associated communities were found. Seawater communities were
mainly comprised of the OM60 (NOR5) clade, \emph{Litoricola},
\emph{Acinetobacter} and the SAR86 clade, while epiphytic communities
were mainly composed of no relative \emph{Gammaproteobacteria} and
\emph{Granulosicocus}. Beside these two groups specific to all three
epiphytic communities, \emph{C. nodosa} was characterized by
\emph{Arenicella}, no relative \emph{Burkholderiales} and
\emph{Methylotenera}, while \emph{Thioploca}, no relative
\emph{Cellvibrionaceae} and \emph{Reinekea} were more specific to both
invaded and noninvaded \emph{C. cylindracea}. In addition,
\emph{Arenicella} was more pronounced in November and December 2017,
while no relative \emph{Burkholderiales} and \emph{Methylotenera} were
more characteristic for the period form March until May 2018. For the
\emph{C. cylindracea} specific taxa no relative \emph{Cellvibrionaceae}
and \emph{Reinekea} showed some seasonality and were characterisitic for
samples originating from June to October 2018. In addition, similarly to
\emph{Bacteroidota}, a large proportion of the surface associated
community was grouped as other \emph{Gammaproteobacteria} indicating
high diversity within this group (\autoref{gamma}).

@@

Seasonal richness changes in the epiphytic community were substantial as
could be observed in the proportion of OTUs that could be find at every
sampling time (\(\leq\) 1.0 \si{\percent}). Interestingly, these OTUs
were accounting for a high proportion of sequences (\(\leq\) 52.8
\si{\percent}). A very similar proportion of persistent OTUs and their
sequence contribution was reported in high frequency studies describing
seasonal picoplankton changes (Gilbert \emph{et al.}, 2009, 2012). In
comparison to the seawater community, a lower degree of seasonal shifts
was observed for the surface associated communities. It seems, that
microniches on the surfaces of macrophytes are providing more stable
conditions in comparison to the seawater. At the level of OTUs seasonal
changes of \emph{C. nodosa} and \emph{C. cylindracea} associated
communities were identified that could be linked to the growth cycle of
the seagrass and macroalgae (M. Najdek, personal communication).
\emph{C. nodosa} was characterized by a Spring community during maximum
seagrass proliferation, a Summer community during a biomass maximum and
a Autumn/Winter community during a biomass decay. In contrast, \emph{C.
cylindracea} started to proliferate in late Spring and was characterized
only by a Summer community during maximal biomass increase and by a
Autumn/Winter/Spring community when the biomass were at the peak and the
settlement started to subsequently decay. Similar seasonal changes in
the epiphytic community was described also for other macroalgae (Tujula
\emph{et al.}, 2010; Lachnit \emph{et al.}, 2011). Higher temporal
stability of \emph{C. cylindracea} surface communities in comparison to
\emph{C. nodosa} were also observed in the higher proportion of shared
communities between two consecutive sampling points.

@@

Illumina MiSeq sequencing of the V4 16S rRNA region was performed as
described previously (Korlević \emph{et al.}, submitted). The V4 region
of the 16S rRNA gene was amplified using a two-step PCR procedure. In
the first PCR the 515F (5'-GTGYCAGCMGCCGCGGTAA-3') and 806R
(5'-GGACTACNVGGGTWTCTAAT-3') primers from the Earth Microbiome Project
(\url{http://press.igsb.anl.gov/earthmicrobiome/protocols-and-standards/16s/})
were used (Caporaso \emph{et al.}, 2012; Apprill \emph{et al.}, 2015;
Parada \emph{et al.}, 2016). These primers contained on their 5' end a
tagged sequence. Purified PCR products were sent for Illumina MiSeq
sequencing at IMGM Laboratories, Martinsried, Germany. Before sequencing
at IMGM, the second PCR amplification of the two-step PCR procedure was
performed using primers targeting the tagged region incorporated in the
first PCR. In addition, these primers contained adapter and
sample-specific index sequences. Beside samples, a positive and negative
control for each sequencing batch was sequenced. Negative control was
comprised of PCR reactions without DNA template, while for a positive
control a mock community composed of evenly mixed DNA material
originating from 20 bacterial strains (ATCC MSA-1002, ATCC, USA) was
used. The sequences obtained in this study have been submitted to the
European Nucleotide Archive (ENA) under accession numbers \textbf{TO BE
ADDED LATER!}.

>>>>

Sequencing of the 16S rRNA V4 region yielded a total of 1.8 million
sequences after quality curation and exclusion of eukaryotic,
chloroplast, mitochondrial and no relative sequences
(\autoref{supp-nseq_notus}). A total of 35 samples originating from
epiphytic archaeal and bacterial communities associated with surfaces of
the seagrass \emph{C. nodosa} and macroalga \emph{C. cylindracea} were
analyzed. In addition, 18 samples (one of the samples was sequenced two
times) originating from picoplankton archaeal and bacterial communities
in the ambient seawater were also processed for comparison. The number
of reads per sample ranged between 8,408 and 77,465 sequences
(\autoref{supp-nseq_notus}). Even when the highest sequencing effort was
applied the rarefaction curves did not level off that is a common
observation in high-throughput 16S rRNA amplicon sequencing approaches
(\autoref{supp-rarefaction}). Following quality curation and exclusion
of sequences mentioned before reads were clustered into 28,736 different
OTUs at a similarity level of 97 \si{\percent}. Reads numbers were
normalized to the minimum number of sequences, 8,408
(\autoref{supp-nseq_notus}), through rarefaction resulting in 17,114
different OTUs that ranged from 379 to 2,016 OTUs per sample
(\autoref{supp-calculators}). To determine seasonal changes of richness
and diversity the Observed Number of OTUs, Chao1, ACE, Exponential
Shannon (Jost, 2006) and Inverse Simpson were calculated after
normalization through rarefaction. Generally, richness estimators and
diversity indices showed similar trends. On average, higher values were
found for \emph{C. cylindracea} (invaded {[}Number of OTUs, 1,676.9 ±
140.3 OTUs{]} and noninvaded {[}Number of OTUs, 1,745.9 ± 156.3
OTUs{]}), middle vales for \emph{C. nodosa} (Number of OTUs, 1,059.5 ±
212.3 OTUs) and lower values for picoplankton communities in the ambient
seawater (Number of OTUs, 530.8 ± 143.1 OTUs)
(\autoref{supp-calculators}). Seasonal changes did not show such large
dissimilarities. \emph{C. nodosa} communities showed a slow increase
towards the end of the study, while \emph{C. cylindracea} (invaded and
noninvaded) communities were characterized by slightly larger values in
Spring and Summer in comparison to Autumn and Winter
(\autoref{supp-calculators}).

To determine the proportion of shared archaeal and bacterial OTUs and
communities sampled in different environments the Jaccard's Similarity
Coefficient on presence-absence data and Bray-Curtis Similarity
Coefficient were, respectively, calculated. Coefficients were determined
after normalization through rarefaction and binning of samples from a
particular environment. The highest proportion of shared OTUs and
community was found between invaded and noninvaded \emph{C. cylindracea}
(Jaccard, 0.35; Bray-Curtis, 0.77), while lower shared values were
calculated between seawater and epiphytic communities
(\autoref{matrix}). Shared proportion between \emph{C. nodosa} and
\emph{C. cylindracea} were approximately in the middle between these two
extremes. To assess seasonal changes in the proportion of shared OTUs
and communities the Jaccard's and Bray-Curtis Similarity Coefficients
were calculated between consecutive sampling points (\autoref{shared}).
Both coefficients showed similar trends. Temporal proportional changes
were more pronounced for seawater in comparison to \emph{C. nodosa} and
especially \emph{C. cylindracea} associated communities
(\autoref{shared}). In addition, only 0.4 -- 1.0 \si{\percent} of OTUs
from each surface associated community were found at every time point.
These OTUs made also a high proportion of total sequences (40.3 -- 52.4
\si{\percent}). To further disentangle the environmental and seasonal
community dissimilarity a Principal Coordinates Analysis (PCoA) based on
Bray-Curtis distances and OTU abundances was applied. It showed a clear
separation between planktonic and surface associated communities
(\autoref{pcoa}). In addition, a separation of epiphytic bacterial and
archaeal communities based on host species was determined. This
separation was further supported by ANOSIM (R = 0.96, \emph{p}
\textless{} 0.001). Seasonal changes of \emph{C. nodosa} associated
communities indicated a separation between Spring, Summer and
Autumn/Winter samples (ANOSIM, R = 0.55, \emph{p} \textless{} 0.001),
while communities from the surfaces of \emph{C. cylindracea} indicated a
non so strongly supported, as in previous cases, separation between
Summer and Autumn/Winter/Spring samples (ANOSIM, R = 0.31, \emph{p}
\textless{} 0.01) (\autoref{pcoa}).

@@

\emph{Cyanobacteria} related sequences were comprising, on average, 5.5
± 4.4 \si{\percent} of total sequences (\autoref{cyano}). Higher
proportions were found for \emph{C. nodosa} (16.4 ± 5.3 \si{\percent})
and \emph{C. cylindracea} (invaded {[} (7.7 ± 3.9 \si{\percent}){]} and
noninvaded {[} (7.8 ± 2.4 \si{\percent}){]}) associated communities in
autumn and for seawater communities in winter (8.8 ± 7.5 \si{\percent}).
Large taxonomic differences between surface associated and seawater
cyanobacterial communities were observed. Seawater communities were
mainly comprised of \emph{Cyanobium} and \emph{Synechococcus}, while
surface associated communities were consisted of \emph{Pleurocapsa} and
sequences without known relatives within \emph{Cyanobacteriia}
(\autoref{cyano}). In addition, seasonal changes in surface associated
communities were observed with \emph{Pleurocapsa} and no relative
\emph{Cyanobacteriia} comprising larger proportions in autumn and winter
and \emph{Acrophormium}, \emph{Phormidesmis} and no relative
\emph{Nodosilineaceae} in spring and summer (\autoref{cyano}).

@@

Sequences related to \emph{Gammaproteobacteria} were comprising, on
average, 18.6 ± 3.9 \si{\percent} of all sequences (\autoref{gamma}).
Similarly to previous taxa, large taxonomic differences between seawater
and surface associated communities were found. Seawater communities were
mainly comprised of the OM60 (NOR5) clade, \emph{Litoricola},
\emph{Acinetobacter} and the SAR86 clade, while epiphytic communities
were mainly composed of no relative \emph{Gammaproteobacteria} and
\emph{Granulosicocus}. Beside these two groups specific to all three
epiphytic communities, \emph{C. nodosa} was characterized by
\emph{Arenicella}, no relative \emph{Burkholderiales} and
\emph{Methylotenera}, while \emph{Thioploca}, no relative
\emph{Cellvibrionaceae} and \emph{Reinekea} were more specific to both
invaded and noninvaded \emph{C. cylindracea}. In addition,
\emph{Arenicella} was more pronounced in November and December 2017,
while no relative \emph{Burkholderiales} and \emph{Methylotenera} were
more characteristic for the period form March until May 2018. For the
\emph{C. cylindracea} specific taxa no relative \emph{Cellvibrionaceae}
and \emph{Reinekea} showed some seasonality and were characterisitic for
samples originating from June to October 2018. In addition, similarly to
\emph{Bacteroidota}, a large proportion of the surface associated
community was grouped as other \emph{Gammaproteobacteria} indicating
high diversity within this group (\autoref{gamma}).

@@

Seasonal richness changes in the epiphytic community were substantial as
could be observed in the proportion of OTUs that could be find at every
sampling time (\(\leq\) 1.0 \si{\percent}). Interestingly, these OTUs
were accounting for a high proportion of sequences (\(\leq\) 52.4
\si{\percent}). A very similar proportion of persistent OTUs and their
sequence contribution was reported in high frequency studies describing
seasonal picoplankton changes (Gilbert \emph{et al.}, 2009, 2012). In
comparison to the seawater community, a lower degree of seasonal shifts
was observed for the surface associated communities. It seems, that
microniches on the surfaces of macrophytes are providing more stable
conditions in comparison to the seawater. At the level of OTUs seasonal
changes of \emph{C. nodosa} and \emph{C. cylindracea} associated
communities were identified that could be linked to the growth cycle of
the seagrass and macroalgae (M. Najdek, personal communication).
\emph{C. nodosa} was characterized by a Spring community during maximum
seagrass proliferation, a Summer community during a biomass maximum and
a Autumn/Winter community during a biomass decay. In contrast, \emph{C.
cylindracea} started to proliferate in late Spring and was characterized
only by a Summer community during maximal biomass increase and by a
Autumn/Winter/Spring community when the biomass were at the peak and the
settlement started to subsequently decay. Similar seasonal changes in
the epiphytic community was described also for other macroalgae (Tujula
\emph{et al.}, 2010; Lachnit \emph{et al.}, 2011). Higher temporal
stability of \emph{C. cylindracea} surface communities in comparison to
\emph{C. nodosa} were also observed in the higher proportion of shared
communities between two consecutive sampling points.

@@

Illumina MiSeq sequencing of the V4 16S rRNA region was performed as
described previously (Korlević \emph{et al.}, submitted). The V4 region
of the 16S rRNA gene was amplified using a two-step PCR procedure. In
the first PCR the 515F (5'-GTGYCAGCMGCCGCGGTAA-3') and 806R
(5'-GGACTACNVGGGTWTCTAAT-3') primers from the Earth Microbiome Project
(\url{http://press.igsb.anl.gov/earthmicrobiome/protocols-and-standards/16s/})
were used (Caporaso \emph{et al.}, 2012; Apprill \emph{et al.}, 2015;
Parada \emph{et al.}, 2016). These primers contained on their 5' end a
tagged sequence. Purified PCR products were sent for Illumina MiSeq
sequencing at IMGM Laboratories, Martinsried, Germany. Before sequencing
at IMGM, the second PCR amplification of the two-step PCR procedure was
performed using primers targeting the tagged region incorporated in the
first PCR. In addition, these primers contained adapter and
sample-specific index sequences. Beside samples, a positive and negative
control for each sequencing batch was sequenced. Negative control was
comprised of PCR reactions without DNA template, while for a positive
control a mock community composed of evenly mixed DNA material
originating from 20 bacterial strains (ATCC MSA-1002, ATCC, USA) was
used. Sequences obtained in this study have been deposited in the
European Nucleotide Archive (ENA) at EMBL-EBI under accession number
PRJEB37267.
